\section{The Problem}
\label{sec:problem}

The applications that dominate the modern web --- group chat, social feeds, collaborative documents, identity, reputation --- run on a small number of centralized service providers. Communication, computation, and storage all flow through their infrastructure. So does the moderation, the surveillance, and the de-platforming risk. The architectural question this paper addresses is how to recover the application-level functionality of those services without recreating the central choke points.

Two broad families of decentralized systems have been deployed at meaningful scale, and neither is a good substrate for interactive applications.

\paragraph{Content-addressed storage.} The original Freenet \citep{clarke2001freenet} and IPFS \citep{benet2014ipfs} provide name-by-hash access to immutable data. Mutable state is whatever the application chooses to layer on top. This is a poor fit for any system whose primary action is mutation: chat rooms, document editing, social feeds. In practice the application reinvents a consistency story per use case, often inconsistently across clients.

\paragraph{Blockchains.} Ethereum \citep{wood2014ethereum} and successor smart-contract platforms provide programmable mutable state, but their consistency model is global linearizable consensus. Every state transition is ordered with every other transition, and replicated by every participating node. The latency, throughput, and replication cost that follow are reasonable for value-transfer and authoritative ledgers, but they are too expensive for the interactive applications that constitute the bulk of the web.

The gap between these is large. A chat room does not need a global linear order on every message; it needs replicas that converge to the same content. A reputation system does not need every reader to download every signed record; it needs efficient summary-driven sync. A collaborative document does not need consensus on each keystroke; it needs an algebra of merge that the application defines.

This paper describes a peer-to-peer platform built around that observation. Applications carry their own merge semantics, expressed as sandboxed WebAssembly. The platform handles routing, replication, synchronization, transport, and sandboxing generically over whatever algebra the application supplies, alongside a separate primitive --- the delegate --- for private state and sensitive operations that should not be replicated. The result targets a class of decentralized applications that has historically been awkward to build on either content-addressed or blockchain substrates.
