\section{Design Thesis}
\label{sec:thesis}

The platform's central bet is to separate \emph{what merges} from \emph{how it propagates}. The application defines the algebra of its state: an idempotent, commutative merge operation on byte-string state, packaged as a WebAssembly module. The platform defines the rest: where state lives on the network, how peers find each other, how they exchange differences efficiently, how updates fan out to subscribers, how secrets are held locally. The platform is generic over the algebra; the application carries the lattice.

This design choice is what distinguishes the system from both ends of the existing decentralized spectrum.

\paragraph{Against fixed-lattice CRDTs.} Conventional Conflict-free Replicated Data Type \citep{shapiro2011crdt} frameworks expose a closed catalogue of lattices (counters, sets, registers, sequences) to applications. Each application chooses from the menu. The platform here generalizes the menu: a contract supplies its own merge function, with the validity invariants and tiebreak rules its workload requires. The trade is that the runtime cannot statically verify the algebraic properties: a contract whose ``merge'' is not associative will not converge. The contract is content-addressed, however, so any deviation is attributable to a specific module that users can choose to use or avoid.

\paragraph{Against global consensus.} Blockchain platforms order all state transitions globally. That is more than most decentralized applications need. The system here provides per-contract eventual consistency: each contract's state converges across its own replicas without requiring agreement with the rest of the network on the order in which its updates were applied. This makes the interactive, conversational class of applications natural, at the deliberate cost of being unsuitable for applications that require a single global order.

\paragraph{Public state and private state are different problems.} Real applications have both: a chat room is public, the user's signing key is private. The platform makes the distinction first-class. Contracts hold the public, replicated state. Delegates hold the private, device-local state. A clean interface between them, with attributed messages crossing a platform-enforced trust boundary, replaces the usual story of ``each application figures out where to keep its secrets,'' which has not worked well in practice.

The remainder of the paper describes the primitives that follow from this thesis (\cref{sec:primitives}), how updates move through the network (\cref{sec:updates}), how data is located (\cref{sec:routing}), how trust is established and what it does not cover (\cref{sec:trust}), the current implementation status and open problems (\cref{sec:status}), and the position in related work (\cref{sec:related}).
