\section{Contracts as Commutative Monoids}
\label{sec:monoids}

The central abstraction in Freenet is the requirement that every contract equip its state space with the structure of a \emph{commutative monoid}. This is what makes eventual consistency a property of the network rather than of each application's hand-rolled protocol.

\subsection{Definitions}

\begin{definition}[Contract state algebra]
\label{def:algebra}
A contract $\contract$ defines:
\begin{enumerate}[leftmargin=*, itemsep=2pt, topsep=2pt]
  \item a \emph{state space} $\statespace_\contract \subseteq \{0,1\}^*$ of byte strings together with a \emph{validity predicate} $\valid_\contract : \{0,1\}^* \to \{\bot, \top\}$;
  \item an \emph{identity element} $e_\contract \in \statespace_\contract$ with $\valid_\contract(e_\contract) = \top$;
  \item a binary \emph{merge} operation $\merge_\contract : \statespace_\contract \times \statespace_\contract \to \statespace_\contract$.
\end{enumerate}
The triple $(\statespace_\contract, \merge_\contract, e_\contract)$ is required to be a \emph{commutative monoid}: for all $a, b, c \in \statespace_\contract$,
\begin{align*}
  a \merge_\contract (b \merge_\contract c) &= (a \merge_\contract b) \merge_\contract c &&\text{(associativity)}\\
  a \merge_\contract b &= b \merge_\contract a &&\text{(commutativity)}\\
  a \merge_\contract e_\contract &= a &&\text{(identity)}.
\end{align*}
We further require $\merge_\contract$ to be \emph{idempotent}: $a \merge_\contract a = a$.
\end{definition}

A commutative \emph{and} idempotent monoid is equivalently a join-semilattice; this is precisely the algebraic structure of a state-based Conflict-free Replicated Data Type \citep{shapiro2011crdt}. We choose the monoid presentation because it foregrounds the operational property the platform actually relies on: arbitrary peers can merge each other's state at any time, in any order, with any multiplicity, and arrive at the same result.

The validity predicate $\valid_\contract$ is checked at the boundary: a peer rejects a candidate state that fails the predicate and never merges it. Validity typically encodes integrity invariants beyond what the type system enforces, for example ``every record in the log carries a valid signature by an authorized key'' or ``the counter is non-decreasing.''

\subsection{Updates}

An \emph{update} is itself an element of $\statespace_\contract$. When a peer holding state $\sigma$ receives an update $u$, it computes the new state as $\sigma' = \sigma \merge_\contract u$, subject to $\valid_\contract(\sigma') = \top$. Application code that produces an update simply produces a contract state that captures the change: a new post for an append-only feed is the singleton-feed state containing only that post; an increment of a counter is the counter-value state containing the new count. There is no separate notion of ``operation'' on a contract; the algebra of state is the algebra of change.

\subsection{Convergence}

The defining property of the algebra is that any two peers that have received the same set of valid update messages, in any order, agree on the resulting state.

\begin{proposition}[Convergence]
\label{prop:convergence}
Let $u_1, \ldots, u_n \in \statespace_\contract$ be a finite multiset of valid updates and let $\sigma_0 \in \statespace_\contract$ be a starting state. For any permutation $\pi$ of $\{1, \ldots, n\}$,
\[
  \sigma_0 \merge_\contract u_{\pi(1)} \merge_\contract \cdots \merge_\contract u_{\pi(n)} = \sigma_0 \merge_\contract u_1 \merge_\contract \cdots \merge_\contract u_n.
\]
\end{proposition}

\noindent \emph{Argument.}\ Immediate from associativity and commutativity of $\merge_\contract$. Idempotence further implies that duplicate delivery of an update is harmless. Together these properties are the standard sufficient conditions for the Strong Eventual Consistency theorem of Shapiro et al.\ \citep{shapiro2011crdt}: peers that have delivered the same set of updates converge regardless of order or repetition.

\subsection{Worked Examples}

\paragraph{Monotone counter.} $\statespace = \mathbb{N}$, $\merge = \max$, $e = 0$, $\valid \equiv \top$.

\paragraph{Last-writer-wins register.} $\statespace = V \times \mathbb{N} \times H$ where $V$ is the value type, the natural number is a Lamport timestamp, and $H$ a tiebreaker hash. $(v_1, t_1, h_1) \merge (v_2, t_2, h_2)$ is the triple with the lexicographically greater $(t, h)$. Identity is $(\bot, 0, 0)$.

\paragraph{Append-only signed log.} $\statespace$ is the set of finite sets of records, each record carrying a signature by an authorized key. $\merge = \cup$, $e = \emptyset$. Validity rejects any state containing a record whose signature does not verify against a key listed in $\mathrm{params}$. A bounded variant evicts the oldest records once a configured size limit is exceeded, while preserving idempotence by making eviction a deterministic function of the merged set.

\paragraph{Observed-remove set.} A set with tombstones: $\statespace = (E, T)$ where $E$ is a set of $(\mathrm{elem}, \mathrm{id})$ pairs and $T$ is a set of $\mathrm{id}$s that have been removed. $\merge$ is pairwise union; the materialized set is $\{\mathrm{elem} : (\mathrm{elem}, \mathrm{id}) \in E,\ \mathrm{id} \notin T\}$.

\subsection{The Generalization over CRDTs}

A conventional CRDT runtime provides a closed catalogue of lattices (G-counter, OR-set, RGA, and so on) and applications choose from the menu. The Freenet contract interface inverts this: the application supplies the lattice in WebAssembly, and the platform is generic over it. Two consequences follow.

First, the platform cannot statically verify that a contract's $\merge_\contract$ is in fact associative, commutative, and idempotent. A contract that fails these properties will exhibit non-converging replicas. Because the contract's code is hashed into its key, however, any divergence is attributable to a specific, content-addressed module: a contract that does not converge is bug-compatible with the rest of the network only insofar as users choose to use it. We treat the algebraic properties as obligations of the contract author, not invariants the runtime enforces.

Second, an application can choose merge semantics tuned to its workload --- a custom CRDT, a domain-specific signed-CRDT variant, a deterministic resolution of conflicting writes by application policy --- without depending on the platform to bless that choice. We have found this flexibility necessary in practice: the consistency requirements of a chat room, a reputation contract, and a public file feed are very different, and bundling all of them into a single lattice family would either force one to pay another's cost or leave several of them unimplementable.
