\section{How Data Is Found}
\label{sec:routing}

A request for contract $\contract$ originates at some peer and must reach a peer that hosts $\contract$'s state. This section describes the ring topology, the rule that constructs it, the routing logic that uses it, and the bootstrap by which a new peer joins. The section closes with a 24-hour measurement of routing path lengths on the live network.

\subsection{Locations and the Ring}

A peer's location $\ell(p)$ is computed by hashing its external network address into $[0,1)$. A contract's location $\ell(\contract)$ is similarly derived from its content-addressed key. Distance is the wrap-around metric of \cref{sec:primitives}.

This is the simplest workable arrangement, and its trade-offs are worth being explicit about. Locations are stable across restarts that preserve the peer's address but change when a peer is rehomed, when NAT mappings shift, or when a peer is reachable on multiple addresses. An adversary in control of many addresses (a cloud subnet, an IPv6 range, a small botnet) can grind locations and concentrate them in chosen regions of the ring. The platform does not attempt to solve Sybil resistance at this layer; defenses are application-level (signed contributions, reputation contracts, capacity-limited delegation) and are discussed in \cref{sec:trust}.

\subsection{Why Small-World Routing}
\label{sec:smallworld}

The platform's routing strategy traces to Kleinberg's small-world result \citep{kleinberg2000small}. Place $N$ peers on a ring (or any metric space). Give each peer a constant number of \emph{short-range} links to its immediate neighbors on the ring, plus a small number of \emph{long-range} links sampled from a distribution that scales as $1/d$ in the ring distance $d$. Kleinberg proved that a node who knows only its own neighbors can route a message to any target in $O(\log^2 N)$ hops by greedy forwarding: at each step, send the message to whichever neighbor is closest to the target. No global knowledge, no flooding, no routing-table maintenance beyond keeping the link distribution roughly $1/d$.

The $1/d$ distribution is what makes the bound work. Too short and the long-range links don't help cover ground; too long and they overshoot the target. $1/d$ produces, at each scale, a constant probability of a useful long-range hop, so a request roughly halves its remaining distance at every step until the local neighborhood dominates.

The practical question is how the $1/d$ distribution gets there. Kleinberg's analysis assumed it as a given. Symphony \citep{manku2003symphony} sampled the long-range links from a harmonic distribution at join time and kept them. The platform here takes a different tack: each peer continuously \emph{steers} its own neighborhood toward the $1/d$ shape, by choosing the destination of each new \textsf{CONNECT} request from the under-represented region of its current connection distribution (\emph{gap-targeting}, \cref{sec:routing}), and by preferring incoming connection candidates that close large gaps in its existing distribution (\emph{Kleinberg-style score}, also \cref{sec:routing}). The active steering is what lets the network preserve the $1/d$ shape under churn; this is the mechanism we describe in the rest of the section.

\subsection{Topology Construction}

The small-world topology is built by the \textsf{CONNECT} operation. A joiner chooses a \emph{desired location} on the ring, sends a \textsf{CONNECT} message routed greedily toward that location, and the peer that terminates the route decides whether to accept. Both ends of this exchange are more involved than the original ``accept only at the terminus'' framing suggests, and both are actively shaping the topology toward Kleinberg's $1/d$ ideal \citep{kleinberg2000small} rather than relying on it to emerge.

\input{figures/ring-gap.tex}

\paragraph{Desired location.} A peer with fewer than $K_0$ existing connections targets its own location, with random jitter applied after consecutive acceptance failures so that subsequent attempts probe different ring regions. Joiner locations are derived from network-address prefixes (\cref{sec:trust}), so distinct joiners hit distinct early targets and bootstrap traffic naturally spreads across the gateway's neighborhood. A peer with at least $K_0$ connections instead computes a \emph{gap target}: it measures the ring distance from itself to each existing neighbor, identifies the largest gap in that distribution in log-distance space, and targets the gap's midpoint. Gap-targeting is the active mechanism by which each peer pulls its own connection distribution toward $1/d$ as it accumulates neighbors: the locations under-represented by current connections are exactly the locations the next \textsf{CONNECT} targets. The deployed thresholds are $K_0 = 3$ for the initial-join path and $K_0 = 5$ for the steady-state maintenance loop; the slightly higher maintenance threshold lets the local connection distribution settle before refresh \textsf{CONNECT}s start gap-targeting.

\paragraph{Acceptance.} A peer that receives a \textsf{CONNECT} applies a chain of decisions. If it can forward strictly closer to the target, it forwards; this is the original terminus rule. A peer within a small fraction of the ring of the target (the deployed band is 5\%) may additionally accept with a proximity-scaled probability while still forwarding, giving the joiner more than one nearby connection per request and softening the strict-terminus discipline. When a request reaches its terminus, the receiver's choice depends on its current connection count: well below a small threshold it accepts unconditionally to unblock bootstrap; below the configured minimum-connections target it filters by a Kleinberg-style score on the candidate's location relative to its existing neighbors, with the acceptance floor sliding from generous (early) to selective (near full); at or above the minimum it admits only the best-scoring candidate from a rolling window of recent arrivals. Above the maximum it rejects. When the terminal receiver rejects, the \textsf{CONNECT} may also be re-forwarded \emph{uphill} (away from the target, up to a small budget of additional hops, deployed value 8) so that a joiner whose precise target lies behind a saturated peer can still be picked up by a nearby non-saturated one.

\input{figures/connect-path.tex}

\paragraph{Where the long-range edges come from.} The long-range component of the connection distribution is the joint output of gap-targeting and Kleinberg-score acceptance, not a side effect of ordinary request traffic. A peer with mostly-local neighbors has its largest log-distance gap at long range, so the next \textsf{CONNECT} it issues lands at a long-range terminus. The receiver, in turn, prefers candidates whose locations close large gaps in its own neighbor distribution. The combination pulls each peer's neighborhood toward $1/d$ from both ends.

\paragraph{Bounded fan-out.} Admission control bounds the connection count between configurable minimum and maximum (deployed defaults 25 and 200).

\subsection{Greedy Forwarding}

A request from peer $p$ to target location $t$ proceeds as follows. Let $N(p)$ be $p$'s current set of neighbors. If some neighbor $q \in N(p)$ is strictly closer to $t$ on the ring than $p$ itself, forward the request to the closest such $q$; otherwise $p$ is a terminus for the request and serves it locally.

Each request carries a \emph{hops-to-live} (HTL) counter that decrements on each forward. The deployed maximum is 10. Above a configurable threshold (HTL $> 7$ by default) the next hop is chosen by random walk rather than greedily. The fallback ensures the request leaves the immediate neighborhood of a peer that happens to be very near $t$ but has not cached the contract yet, and contributes path diversity that helps the topology mix.

\subsection{Adaptive Routing}

Greedy-by-distance assumes the topologically closest neighbor is also the operationally fastest path. In a real peer-to-peer network this is wrong often enough to matter: neighbors vary by an order of magnitude or more in bandwidth, RTT, and uptime. The router therefore picks the next hop not by ring distance alone but by predicted total cost.

For each neighbor $q$, the local peer maintains a history of routing events of the form $(d, \tau, b, f)$, where $d$ is the ring distance from $q$ to the request's target, $\tau$ is the observed time-to-first-byte, $b$ is the observed transfer rate, and $f$ is a failure indicator. From this history three estimators are fit:
\[
  \widehat{\tau}_q(d), \quad \widehat{p}_q(d), \quad \widehat{b}_q(d),
\]
the predicted latency, failure probability, and throughput as functions of distance-to-target. Each is fit by \emph{isotonic regression} \citep{barlow1972statistical, deleeuw2009isotone}, the standard non-parametric estimator for a monotone function.

The choice of isotonic regression encodes a structural prior we are willing to commit to (latency and failure are non-decreasing in distance-to-target, on average) without committing to a parametric family for the shape of the curve. The pool-adjacent-violators algorithm fits each curve in time logarithmic in history size; old events age out so the model tracks current conditions.

Given a target $t$ for a new request, the router selects the neighbor that minimizes an expected-cost score combining the three estimators. A neighbor with too little history falls back to greedy-by-distance until enough events have accumulated. Predictions between fitted points use linear interpolation, and beyond the training-data domain the model extrapolates linearly.

\paragraph{Where the adaptive layer applies.} For ordinary request traffic (\textsf{GET}, \textsf{PUT}, \textsf{SUBSCRIBE}, \textsf{UPDATE}) the adaptive estimator \emph{replaces} the greedy rule of the previous subsection once the local model has enough data: the router picks by predicted cost, not by ring distance. \textsf{CONNECT} is different. \textsf{CONNECT} additionally retains a hard greedy filter (only neighbors strictly closer to the target than the current peer are eligible) and uses a separate isotonic estimator trained on Bernoulli accept/reject outcomes to rank among those eligible candidates. The hard greedy filter keeps joiner traffic moving downhill so the topology-construction logic in \cref{sec:routing} converges; ordinary requests do not need that guarantee and benefit from the broader candidate set.

\input{figures/adaptive-routing.tex}

\subsection{Lifecycles}

The five operations of \cref{sec:primitives} are realized as transaction state machines coordinated by the local core. Each is assigned a transaction id, tracked through completion, and supports timeout, retry, and composite parent/child relationships.

\paragraph{\textsf{CONNECT}.} As above. A successful \textsf{CONNECT} establishes a single neighbor relationship. A joining peer issues an initial sequence of \textsf{CONNECT} operations to populate its first $K_0$ neighbors using own-location targeting; once it has enough connections for gap analysis, all subsequent \textsf{CONNECT} operations use gap-target locations. Topology growth and refresh are driven by additional \textsf{CONNECT} operations issued from a maintenance loop, not as a side effect of ordinary request traffic.

\paragraph{\textsf{PUT}.} A \textsf{PUT} carries a contract $\contract$ and an initial state $\sigma_0$. The originating peer routes the request toward $\ell(\contract)$. The first peer with capacity at or near $\ell(\contract)$ caches $\contract$, checks $\valid_\contract(\sigma_0)$, and accepts the publish. Nearby peers receive the state through the same path and replicate it.

\paragraph{\textsf{GET}.} A \textsf{GET} carries a key $k$. The originating peer routes toward the corresponding location. Any peer along the path that currently hosts the contract may answer with its current state $\sigma$ and parameters; the response travels back along the request path, optionally caching at intermediate peers.

The requester can verify on arrival that the served code and parameters hash to $k$. \emph{State} is not directly verifiable from the key: the key is derived from $\mathrm{code}$ and $\mathrm{params}$, not from state, and state is mutable. This is a meaningful difference from immutable content-addressed stores such as IPFS, where $\mathsf{get}(k)$ returns a value whose own hash equals $k$. Readers carrying that mental model should note that the guarantee does not transfer to Freenet contracts. State is trusted only to the extent that the contract's validity predicate accepts it; replication across several peers near $\ell(\contract)$ is the structural defense against a single peer serving a stale or doctored state.

\paragraph{\textsf{UPDATE}.} An \textsf{UPDATE} carries a key $k$ and a candidate state $u \in \statespace_\contract$. It is routed toward $\ell(\contract)$. Each peer along the path that already hosts the contract merges $u$ into its local copy subject to the validity check, then propagates the change outward through the subscription tree (\cref{sec:updates}).

\paragraph{\textsf{SUBSCRIBE}.} A \textsf{SUBSCRIBE} establishes a lease on future updates. The subscriber routes the request toward $\ell(\contract)$ and the first hosting peer responds; intermediate peers along the path record the subscription. Subscriptions are leased and renewed periodically (\cref{sec:updates}).

\subsection{A 24-Hour Snapshot of Routing on the Deployed Network}
\label{sec:routing-measurement}

\input{figures/connection-distance-histogram.tex}

This subsection records a single observation of routing behaviour on the deployed network, as a sanity check rather than as a scaling study. We also include a snapshot of the connection-distance distribution (\cref{fig:connection-distance}), which is the most direct empirical check on the small-world topology the rest of this section constructs. The 24-hour window ends 2026-05-24. Network size in that window was 443 distinct active peers in the last 10 minutes, with 604 distinct peer identifiers seen across the full 24 hours (the difference attributable to churn). All peers run with telemetry enabled in the current deployment.

For each \textsf{GET} and \textsf{PUT} transaction we compute a \emph{path length}: the number of distinct peers seen emitting the corresponding \textsf{*\_request} event for that transaction id, which is the number of peers visited (hops $+ 1$). Statistics over operations whose telemetry reached completion in the window:\footnote{\textsf{UPDATE} (median path length 2, $n = 386$) and \textsf{SUBSCRIBE} (median 1, $n = 5{,}359$) propagate primarily through pre-existing subscription trees rather than fresh greedy routes, so their path lengths reflect tree depth from the subscriber to the nearest hosting peer, not routing-layer hop counts. They are excluded from the table here.}

\begin{center}
\small
\begin{tabular}{lrrrrr}
\toprule
Operation & $n$ & mean & median & p95 & p99 \\
\midrule
\textsf{GET} & 1{,}302 & 6.84 & 7 & 11* & 11* \\
\textsf{PUT} &    158 & 6.01 & 7 & 10* & 10* \\
\bottomrule
\end{tabular}
\end{center}

\noindent\textit{*The deployed hops-to-live cap is 10, so any route that would otherwise visit 12 or more peers is dropped before it can appear in the table. The reported p95 and p99 for \textsf{GET} are therefore at the cap itself; they record where the tail is truncated, not where it actually lives. We are not directly observing the upper tail of the path-length distribution.}

Two further caveats are worth being explicit about. First, this is one observation at one network size. At $N \approx 443$, $\log_2 N \approx 8.8$; the observed median path length of 7 peers (6 hops) for \textsf{GET} is consistent with logarithmic-diameter routing, but a single data point at a single $N$ cannot distinguish $O(\log n)$ from $O(\sqrt{n})$, $O(n^{0.4})$, or a constant. Second, the table does not compare against any baseline: it does not separate the contribution of the adaptive-routing layer (\cref{sec:routing}) from baseline greedy-by-distance, and it does not measure what the path length would be without the random-walk fallback at high HTL.

A real test of the routing claims is a controlled simulation sweep across $N$ using the deterministic harness described in \cref{sec:status}, with and without the adaptive and random-walk components. We treat that sweep as the appropriate follow-up empirical work; it is listed among the open problems in \cref{sec:status}.
