\section{Implementation Status and Open Problems}
\label{sec:status}

This section describes the reference implementation, makes the implementation status of each major mechanism explicit, and lists the architectural questions we view as open.

\subsection{Implementation}

The Freenet Core is implemented in Rust as an asynchronous \texttt{tokio}-based runtime. Contracts and delegates execute inside Wasmtime \citep{wasmtime}; both run under per-instance fuel limits and explicit memory bounds so that a runaway or malicious module cannot exhaust the host. Contract state is persisted through a pluggable store layer with implementations on top of \texttt{redb} \citep{redb} and SQLite.

The runtime is organized as a single central event loop, a \texttt{tokio::select!} over inbound transport packets, client API events, operation timers, and background tasks, that dispatches into per-operation state machines. The single-coordinator design simplifies reasoning about partial state but limits the throughput of a single peer to what one event-loop thread can sustain. So far this has not been the binding constraint in deployment.

\paragraph{Deterministic simulation testing.} A peer-to-peer system is hard to test by ordinary means: many bugs in routing, congestion, or convergence appear only in specific multi-peer interleavings that are difficult to reproduce. The Core ships with a deterministic simulation harness built on Turmoil \citep{turmoil}: virtual time, a seeded global RNG, an in-memory socket implementation, and fault-injection knobs for loss, latency, partitions, and reorderings. Tests run the entire network in a single deterministic process and replay bit-for-bit from a seed. Most non-trivial routing, subscription, and contract-execution bugs we have found were first reproduced in simulation.

\paragraph{Content-addressed components.} Contracts, delegates, and UI bundles are all addressed by hash. This pushes a lot of complexity off the platform: there is no version negotiation, no upgrade protocol, no need for the platform to understand application semantics during deployment. A new version is published under a new key; the old key continues to work as long as anyone hosts it.

\subsection{Implementation Status of Each Mechanism}

The following table is the current honest status of the mechanisms described in this paper. ``Deployed'' means present in the production runtime and exercised on the live network; ``Implemented'' means present in the runtime but not yet broadly exercised at scale; ``Experimental'' means available as a build-time option but not in production; ``Open'' means designed or partially designed but not implemented.

\begin{center}
\small
\begin{tabular}{p{0.42\textwidth}p{0.18\textwidth}p{0.32\textwidth}}
\toprule
Mechanism & Status & Notes \\
\midrule
Small-world ring topology & Deployed & Live network $\approx 440$ active peers \\
Greedy forwarding + HTL fallback & Deployed & HTL cap = 10, random walk above 7 \\
Two-stage CONNECT acceptance & Deployed & Greedy forward, accept-and-forward in 5\% band, Kleinberg-scored at terminus \\
Adaptive isotonic-regression routing & Deployed & Per-neighbor distance-indexed model \\
Summary/delta synchronization & Deployed & Per-contract, application-defined \\
Subscription trees with leases & Deployed & 8-min lease, 2-min renewal \\
Contract WebAssembly execution & Deployed & Wasmtime, fuel + memory limits \\
Delegates (local, secret state) & Deployed & Used by River chat \citep{river} \\
Delegate cross-device sync via shared-secret contract & Partial & Encrypted-contract primitives exist; end-to-end delegate flow not wired up \\
NAT hole-punching transport & Deployed & UDP, X25519 + AEAD \\
Fixed-rate congestion control & Deployed & Production default \\
LEDBAT++ congestion control & Experimental & Did not meet our requirements in initial trials; not in production \\
BBR-style congestion control & Experimental & Build-time option, not the default \\
Deterministic simulation testing & Deployed & Turmoil-based \citep{turmoil} \\
Per-peer resource limits (bandwidth, storage) & Deployed & User-configured budgets \\
Sybil-resistant peer identity / location & Open & Defenses are application-layer (see \cref{sec:trust}) \\
Network-level incentive / abuse layer & Open & Application-layer reputation contracts are the intended mechanism \\
\bottomrule
\end{tabular}
\end{center}

The single biggest gap between this paper's narrative and the deployed system is congestion control. The platform supports pluggable congestion controllers but the deployed default is a token-bucket fixed-rate scheme. LEDBAT++ and BBR variants exist in the codebase as alternatives, but neither has yet been adapted to the workload satisfactorily. Real congestion control across heterogeneous peer-to-peer links remains open work.

\subsection{Where the Platform Fits}

The combination of contract-defined merge, summary/delta sync, small-world routing, and delegates is well-matched to applications whose consistency requirements are conflict-mergeable in some application-defined sense. Group chat, collaborative documents, social feeds, reputation systems, asynchronous email-like inboxes, and identity directories are all in the well-fit class. A working chat implementation \citep{river} and an anonymous identity scheme \citep{ghostkey} are deployed against the live network.

\subsection{Where the Platform Does Not Fit}

Several application classes do not fit the model gracefully, and we want to be explicit about them.

\paragraph{Strictly linearizable transactions.} A defense against double-spending fungible tokens requires that two attempts to spend the same balance cannot both succeed. The merge model cannot express ``one of these updates is rejected unconditionally''; merge is total. Concretely, fungible-token transfer with platform-level double-spend protection cannot be expressed by a single contract in the current model. Applications that need a strong-consistency primitive must either contain the inconsistency inside a contract whose validity predicate detects and rejects the second update (which forces the conflict into application code, with the loser's state restored on detection) or layer an auxiliary consensus mechanism on top of the platform.

\paragraph{Abuse-resistant open posting.} A purely-open contract is simple to write as a set under union and equally simple to flood. Practical contracts in this class restrict writes to signed records by approved keys, charge a proof-of-work fee per update, or require credentials issued by a reputation contract. The platform's transport-layer rate limiting and per-neighbor resource accounting do provide a partial floor: a neighbor consuming excessive resources is disconnected, which caps the volume any single attacker can push through any single peer. That bound is useful but limited; it does not stop a distributed flood from many cheap identities, nor a slow-burn pattern that stays within per-peer limits. The platform has no application-level spam defense beyond that transport floor.

\paragraph{Coordinated DoS.} Per-peer resource limits protect a peer against a single misbehaving neighbor, not against coordinated traffic from many neighbors using cheap identities. Defenses here probably require accounting that persists across identities, which is the open incentive layer noted above.

\subsection{Open Problems}

We treat the following as genuinely open and intend them as subjects of future work or empirical evaluation.

\begin{enumerate}[leftmargin=1.4em, itemsep=2pt]
  \item \textbf{Scaling characterization.} The single 24-hour data point in \cref{sec:routing-measurement} is consistent with logarithmic-diameter routing at the deployed network size. A real characterization requires scanning $N$ across orders of magnitude. We do not currently have the deployment to do this; controlled simulation studies are an interim alternative.
  \item \textbf{Failure modes of contract-defined merge.} A contract whose merge is not actually associative or idempotent does not converge; the platform's handling of this is currently to allow the divergence to be observable but not to take corrective action. Better diagnostics, and where feasible contract-level invariants checked at the platform layer, are open work.
  \item \textbf{Congestion control.} As above. The current fixed-rate default is a deliberate choice not to ship something we cannot tune correctly, not an endorsement of fixed-rate as the right answer.
  \item \textbf{Sybil-resistant identity.} The application-layer approach has examples (Ghost Keys) but no clean general story; this is the recurrent question for every decentralized platform, and we do not claim to have solved it.
  \item \textbf{Incentive / abuse layer.} How a network as a whole resists coordinated denial of service without a centralized authority, while still allowing low-friction joining, is the work that determines whether the network grows past a small community.
\end{enumerate}
