\section{Related Work}
\label{sec:related}

\paragraph{Distributed hash tables.} Chord \citep{stoica2001chord}, Pastry \citep{rowstron2001pastry}, and Kademlia \citep{maymounkov2002kademlia} established the structured-overlay paradigm the ring inherits. These systems organize peers by hashed identifiers in a ring or XOR-metric space and provide $O(\log n)$ lookup with carefully maintained routing tables. Their consistency model is fundamentally immutable: a put-then-get returns the published value. The platform here replaces the immutable cell with a mutable, contract-defined one and replaces explicit finger-table maintenance with the implicit topology construction of \cref{sec:routing}.

\paragraph{Content-addressed storage.} IPFS \citep{benet2014ipfs} popularized hash-based naming as a deployment substrate for web content, with a peer-to-peer block exchange (Bitswap) underneath. IPFS is also fundamentally immutable, with mutability provided externally (IPNS, smart contracts) and consistency left to applications. The choice here is to put mutability and consistency inside the platform via the contract abstraction; this is strictly more opinionated.

\paragraph{CRDTs.} The theory of Conflict-free Replicated Data Types \citep{shapiro2011crdt} is the algebraic foundation the contract layer builds on. The contract interface realizes the state-based (CvRDT) flavor with two extensions: the lattice is application-supplied rather than chosen from a catalogue, and the summary/delta protocol generalizes the implicit ``ship the whole state'' protocol of vanilla state-based CRDTs. Antidote and AntidoteDB \citep{akkoorath2016cure} are the closest comparison points from the database side, both being CRDT databases for the geo-distributed setting, but they expose a fixed CRDT vocabulary rather than letting applications define their own.

\paragraph{Bayou.} Bayou \citep{terry1995bayou} is the most direct historical precedent for contracts. Application-supplied merge procedures, executed on the replicas, reconciling concurrent writes that the storage layer cannot resolve generically: this is the same idea, in 1995, on trusted replica servers. The differences are forced by the peer-to-peer setting. Bayou's merges were full procedures with database access; contracts are sandboxed WebAssembly subject to the idempotent-commutative-monoid algebra. Bayou's merge identity was per-database configuration; contracts bind merge code to content-addressed keys. Bayou ran on a known set of trusted replicas; contracts run on whatever peers happen to host them. The narrower algebra and the content-addressed binding are the price the platform pays for being able to make a convergence promise without inspecting the merge code or trusting the replica.

\paragraph{Anti-entropy and delta sync.} Dynamo \citep{decandia2007dynamo} and its descendants used Merkle anti-entropy to reconcile replicas. Delta-state CRDTs \citep{almeida2018delta} formalize a state-based protocol that ships only changes between replicas, very much in the same shape as \cref{sec:updates}; the differences here are that the lattice and the delta encoding are both supplied by the contract, and the result is embedded in a small-world routing layer rather than a fully-connected replica set.

\paragraph{Local-first software and CRDT libraries.} The local-first software movement \citep{kleppmann2019local} argues for application-level convergence on user devices over server-mediated consistency. Automerge \citep{automerge} and Y.js \citep{yjs} are the canonical embodiments, offering CRDT data structures (JSON-like trees, sequences, counters) that applications compose into peer-to-peer apps over whatever transport they choose. The platform here shares the local-first goal but inverts the boundary: rather than offering a closed catalogue of CRDT types that applications compose from, it lets each contract supply its own algebra and its own synchronization protocol. The trade is between two design points: library-based CRDT systems offer rigorous convergence for any program built on the library's primitives; contracts let applications go beyond those primitives at the cost of carrying their own correctness obligation.

\paragraph{Secure Scuttlebutt.} SSB \citep{scuttlebutt} is the peer-to-peer system whose primary abstraction is closest in spirit to a single Freenet contract: a per-user append-only signed log. SSB chose one structure (the log) and one replication protocol (gossip on demand by feed identifier); contracts let each application choose. An SSB feed is straightforwardly expressible as a contract whose merge is set-union over signed records, and many SSB applications (private messaging, blogging, social feeds) are recognizably the same applications the contract abstraction targets.

\paragraph{Hypercore.} The Hypercore protocol family \citep{hypercore} provides append-only signed logs replicated peer-to-peer, with higher-level data structures (Hyperbee, Hyperdrive) built on top. As in SSB the abstraction is fixed at the log level; as in contracts the cryptographic identity is content-addressed. The closest analogue to a contract is a Hyperbee whose merge semantics are baked into the library; the closest analogue to a delegate is a local-only key-managed agent acting over Hypercore.

\paragraph{Small-world routing.} Kleinberg's 2000 result \citep{kleinberg2000small} is the theoretical underpinning. The original Freenet \citep{clarke2001freenet} predated and informally anticipated some of it. Symphony \citep{manku2003symphony} introduced explicit sampling from a harmonic distribution to build $1/d$ long-range links on a ring overlay. The platform here uses neither Symphony's static sample nor pure greedy acceptance. Instead it actively converges toward the $1/d$ distribution at each peer: the joiner's \emph{gap target} chooses the under-represented region of the joiner's own connection distribution as the next \textsf{CONNECT} destination, and the receiver applies a Kleinberg-style score filter that prefers candidates closing large gaps in the receiver's neighbor distribution. To our knowledge the specific combination has not been analyzed in the structured-overlay literature.

\paragraph{Performance-aware routing.} Adaptive routing in peer-to-peer overlays has a long history. Isotonic regression specifically has been used as a calibration tool in machine learning \citep{deleeuw2009isotone}; as far as we are aware, it has not previously been applied as the per-neighbor performance model in a structured overlay's forwarding decision.

\paragraph{Blockchain platforms.} Ethereum \citep{wood2014ethereum} and successor smart-contract platforms provide programmable consistency, paid for by a globally-linearized replicated state machine. The platform here relaxes the global linearization in favor of per-contract eventual consistency with application-defined merge. The trade is that a different class of applications is natural: the interactive, conversational ones, at the cost of being unsuitable for applications that fundamentally require a single global transaction order.

\paragraph{Federated systems.} Matrix \citep{matrix} is the closest non-peer-to-peer system in application surface. It is server-based and federated, and its state-resolution algorithm reconciles divergent room states by a deterministic application-defined rule. The platform here pushes the same idea into a fully peer-to-peer setting and generalizes the merge to arbitrary idempotent commutative monoids.

\paragraph{The original Freenet.} The original Freenet \citep{clarke2001freenet} was anonymous, immutable file storage addressed by content-hash keys (CHKs), released around 2000 and described in the 2001 paper. Its routing heuristic (forward to the peer whose previous successful routes most resemble the requested key) was an early precursor to small-world greedy routing, but it operated over a routing table that had no explicit notion of location, and the network's data model was a single immutable blob per key. The platform described here preserves the small-world intuition and the content-addressed naming and replaces almost everything else: explicit locations, contract-defined mutable state, anonymity moved out of the core into application-layer protocols, and first-class delegates and user interfaces alongside contracts. The shared name reflects shared lineage, not shared design.
